---
title: "class 11 alphafold"
author: "Anqi Feng"
format: pdf
---
Today, before delving into structural perdiction with AlphaFold, I will finish off our previous lab10 "comparative structural analysis"

1. sequence identifier
```{r}
library(bio3d)
id <-"1ake_A"
aa<- get.seq(id)
aa
```
```{r}
b <- blast.pdb(aa)
```
```{r}
attributes(b)
```
```{r}
head(b$hit.tbl)
```
```{r}
hits <- plot(b)
```
```{r}
attributes(hits)
```
Top hits we like from our blast results would be 
```{r}
hits$pdb.id
```
```{r}
#Download related pdb files
files<- get.pdb(hits$pdb.id, path = "pdbs", aplit = TRUE, gzip=TRUE)
```
I have now found and downloaded all adk structures in the new database but viewing them is difficult as they need to be aligned and superposed (bioconductor software)

I am going to install BiocManager package from CRAN, then I can use `BiocManager::install()`
to install any bioconductor package
```{r}
#install.packages("BiocManager")
```

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
```{r}
pc<- pca(pdbs)
plot(pc)
```

To examine in more detail what PC1 (or any PC) is capturing here we can plot the loadings or make a wee movie(trajectory) of moving along PC1
```{r}
plot(pc,pc.axes=c(1,2))
```
```{r}
mktrj(pc,pc=1, file = "pc1.pdb")
```



