---
title: "class 5"
author: "Anqi Feng (PID:A16243334)"
format: pdf
---
```{r}
plot(cars)
```

A very popular package in this area is called **ggplot2**.

Before I can use any add-on package like this I must install it with the `install.packages("ggplot2")` command/function

Then to use the package I need to load it with a 
`library(ggplot2)` call. 

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth()
```
for "simple" plots like this one base R code will be much shorter than ggplot
Every ggplot has at least 3 layers
- **data** (data.frame with the numbers and the numbers of stuff you want to plot)
- **aes**thetics (mapping of your columns in a data frame to your plot) 
- **geoms** (there are tones of these, basics, or points: 'geom_line()', 'geom_point()', 'geom_col()')

Q: make me a ggplot of the 'mtcars'data set using 'mpg' vs 'disp'and set the size of the points to the 'hp' and set the color to 'am'.
```{r}
head(mtcars)
library(ggplot2)
ggplot(mtcars) + 
  aes(x=mpg, y=disp, size=hp, col=am) +
  geom_point() 
```
now color all points blue

```{r}
library(ggplot2)
ggplot(mtcars) + 
  aes(x=mpg, y=disp, size=hp) +
  geom_point(color="blue") 
```
```{r}
ggplot(mtcars) + 
  aes(x=mpg, y=disp, size=hp, col = am) +
  geom_point() +
  facet_wrap(~am)
```
```{r}
library(ggrepel)
ggplot(mtcars) + 
  aes(x=mpg, y=disp, col = am, label=rownames(mtcars)) +
  geom_point() +
  facet_wrap(~am)+
  geom_text_repel()
```
```{r}
ggplot(mtcars) + 
  aes(x=mpg, y=disp) +
  geom_point()

ggsave("myplot.png")
```
patchwork is also a good package to install (patches three graphs ex. p1 | p2 | p3) it is a little add on for ggplot

```{r}

library(ggplot2)

url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)
```
There are `r nrow(genes)` genes in this dataset
```{r}

ggplot(genes) + 
    aes(x=Condition1, y=Condition2) +
    geom_point()
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p
p + scale_colour_manual( values=c("blue","gray","red") ) +
    labs(title="Gene Expresion Changes Upon Drug Treatment",
         x="Control (no drug) ",
         y="Drug Treatment")
```
```{r}

#install.packages("dplyr")  ## un-comment to install if needed
library(dplyr)
library(gapminder)
library(ggplot2)
head(gapminder)
gapminder_2007 <- gapminder %>% filter(year==2007)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) + 
  geom_point(alpha=0.4)
```
```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) 
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)cc
```
```{r}

gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5

```
```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop))
```


